Analysis of Programming Students' Prompts to Identify Patterns of Critical Thinking in Interactions with Generative AI \\ {\footnotesize \textsuperscript{*}Note: Sub-titles are not captured in Xplore and should not be used} \thanks{Identify applicable funding agency here. If none, delete this.}

### 2024-08-26 paper/paper.tex
update

\subsection{CharlieBot}

CharlieBot is a ChatGPT-based bot that provides assistance to students in Java
programming courses at UC3M. The bot uses RAG(retrieval augmented generation) to
generate responses to student queries, providing explanations, code snippets,
and debugging help. The goal of CharlieBot is to support students in their
learning process by providing personalized assistance and feedback.

@@

The table \ref{tab:categories} shows the categories of analysis used in this
study. The categories were based on the work of
\cite{Ghimire24} with the addition of four new categories.
Also, we group the categories to understand the impact on critical thinking.

@@

CharlieBot is a ChatGPT-based bot that provides assistance to students in
Java programming courses in UC3M. CharlieBot logs all conversations with
students, which were used for this study. The goal was to categorize student
messages to understand how students interact with the bot and identify patterns
in their study habits.
\\
The categories outlined in \cite{Ghimire24} served as the
foundation for classifying student messages (Debugging Help, Conceptual
Questions, Code Snippet and Complete Solution). However, it was identified that
some categories were missing, necessitating the addition of new categories
(Student Correction, Multiple Question, Language Change and Uncategorized). Table
\ref{tab:categories} presents these categories along with their descriptions.
\\
Subsequently, these eight categories were input into Claude.ai, accompanied by
detailed instructions on the desired structure of the input and the expected
format of the output. For each conversation with the CharlieBot log, a .csv file
was downloaded, and Claude was tasked with classifying the user messages into
one of the eight categories.

The classifications made by Claude were then manually reviewed. In cases where
there was disagreement with the classification, manual adjustments were made,
and the accuracy of Claude's classification was recorded. Each conversation
with CharlieBot was saved into a .csv file containing the original conversation,
along with two additional columns: "Classification" to denote the category and
"AI" to indicate Claude's accuracy.

As new categorizations were completed, an example was submitted to Claude's
knowledge base to enhance accuracy. Initially, Claude began with an example
representing 1\% of the total memory of its knowledge base. As classifications
were progressively completed, more examples were provided to Claude to increase
the percentage of its knowledge base occupied by relevant examples.

Upon completion of all categorizations, all resulting .csv files were consolidated
for final counts and analysis using the Pandas library. This methodology
facilitated an efficient categorization of student messages, thereby enabling
a comprehensive analysis of how tools such as ChatGPT are utilized in foundational
programming courses.

@@

>>>>

\subsection{The CharlieBot}

CharlieBot is a ChatGPT 3.5 based bot that provides assistance to students in a
Java programming courses at UC3M. The bot uses the content of UC3M Java
programming course at EdX to supply a RAG(Retrieval-Augmented Generation) system
and thus try to respond more accurately to students' questions \cite{Sun24}.
The main goal of CharlieBot is support students in their learning process by
providing personalized assistance and feedback.

@@

@@

The study was conducted in two phases: categorization and analysis. The
categorization phase involved classifying student messages into eight
categories. The table \ref{tab:categories} shows the categories of analysis
and their description. The initial categories were based on the work
of \cite{Ghimire24}, however, four new categories were added due to the need to
give more detail to the classification. Beside of that, the categories were
grouped to understand the impact on critical thinking.

The analysis phase involved to understand the way that the students interact
with CharlieBot. The students messages were collected from the logs of the bot
and categorized manually by the authors. A total of YYY messages, of the first
semester of 2024, were analyzed. The participants were students of the YYY
semester of the Computer Science at UC3M where Java programming is taught.

@@

The results of this study suggest that students interact with CharlieBot in a
manner that promotes critical thinking. The majority of the messages analyzed
were related to debugging help, conceptual questions, and student correction,
indicating that students are actively engaging with the bot to understand
concepts and solve problems. This is a positive outcome, as it suggests that
CharlieBot is being used as a tool to enhance learning rather than a crutch for
students to rely on.
